Inferring neural networks from fluorescent calcium imaging using partial correlation and sample weighting

### 2014-06-06 article.tex
link to the challenge website
add todo
move goal to the introduction
remove duplicated content

\begin{abstract}
Understanding how the brain works is a key to understand and to treat
brain pathologies and disorders such as Parkinson's disease, epilepsy.
Retrieving the connectomes, neurons connectivity map, will shed new lights on
the anatomical and functional connectivity of the brain.
In the context of the Connectomics challenge, we propose a pre-processing and a simple approach in order to achieve a high performance inference of the connectivity network.
We put in perspective our method by comparing with other inference method such as GTE [], another method proposed to infer a neural network from calcium fluorescent imaging, and Genie3 [], a tree-based approach very competitive in several gene inference challenge (i.e. DREAM4 and DREAM5).\\

@@

\begin{keywords}
Network inference - Precision matrix - GENIE3
\end{keywords}

@@

% BG : brain, complex organ
Understanding how the brain works is a key to understand and to treat
brain pathologies and disorders such as Parkinson's disease and epilepsy.
Retrieving the connectomes, neurons connectivity map, will shed new lights on
the anatomical and functional connectivity of the brain. However,
the human brain is a complex biological organ, which is formed by 100
billions of neurons with 7000 synaptic connections on average. Inferring the
neuron connectivity network through dissection would be an impossible
and a daunting task.

Cutting edge optical devices [todo cite] allows to track simultaneously
the neural activity of thousands of neurons through a fluorescent
calcium indicator.
From those observational data, the neural connectivity inference consists in
highlighting direct relationships between neurons. However, those imaging
systems have a sampling rate smaller than the neuron firing speed and are
limited to two dimensional image, while the neuronal network is a
three dimensional structure, which may lead to superpose the observed activity
of different neurons. Furthermore, the decay of the fluorescent calcium signal
is slower than its rise.

@@

% Solution
In this paper, we describe a simple and a theoretically grounded approach
based on partial correlation to infer the neural connectivity network.
This is also the winning method of the Connectomics challenge [todo cite].

@@

\section{Network inference methodology}
\label{sec:inference}
% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}
The main challenge in the inference of the network $y$
is to distinguish direct from indirect interactions \cite{de2004discovery}.
For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
from the indirect interaction $A \rightarrow C$.

>>>>

\begin{abstract}
Understanding how the brain works is a key to understand and to treat
brain pathologies and disorders such as Parkinson's disease, epilepsy.
Retrieving the connectomes, neurons connectivity map, will shed new lights on
the anatomical and functional connectivity of the brain.
In the context of the Connectomics challenge, we propose a pre-processing and
a simple approach in order to achieve a high performance inference of the
connectivity network.
We put in perspective our method by comparing with other inference method such
as GTE [], another method proposed to infer a neural network from calcium
fluorescent imaging, and Genie3 [], a tree-based approach very competitive
in several gene inference challenge (i.e. DREAM4 and DREAM5).

@@

\begin{keywords}
Network inference - Partial correlation - Precision matrix - GENIE3
\end{keywords}

@@

% BG : brain, complex organ
% TODO remove redundant information from the abstract
The human brain is a complex biological organ, which is formed by 100
billions of neurons with 7000 synaptic connections on average. Inferring the
neuron connectivity network through dissection would be an impossible
and a daunting task. Cutting edge optical devices [todo cite] allow to track
simultaneously the neural activity of thousands of neurons through a fluorescent
calcium indicator.

From those observational data, the neural connectivity inference consists in
highlighting direct relationships between neurons. However, those imaging
systems have a sampling rate smaller than the neuron firing speed and are
limited to two dimensional image, while the neuronal network is a
three dimensional structure, which may lead to superpose the observed activity
of different neurons. Furthermore, the decay of the fluorescent calcium signal
is slower than its rise.

@@

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}
The main challenge in the inference of the network $y$
is to distinguish direct from indirect interactions \cite{de2004discovery}.
For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
from the indirect interaction $A \rightarrow C$.

% Solution
In this paper, we describe a simple and a theoretically grounded approach
based on partial correlation to infer the neural connectivity network.
This is also the winning method of the Connectomics challenge
\footnote{\url{http://connectomics.chalearn.org/}}.

@@

\section{Network inference methodology}
\label{sec:inference}
